An introduction to Docker for reproducible research, with examples from the R environment

### 2014-09-23 docker-paper/manuscript.Rmd
final edits

It is worth observing from the outset that the primary barrier to
computational reproducibility in many domain sciences has nothing
to do with the technological approaches discussed here, but stems rather from a
reluctance to publish the code used to generating the results in the
first place [@Barnes_2010]. Despite extensive evidence to the contrary
[_e.g._ @Ince_2012], many researchers and journals continue to assume that
summary descriptions or pseudo-code provide a sufficient description of
computational methods used in data gathering, processing, simulation,
visualization, or analysis.  Until such code is available in the first
place, one cannot even begin to encounter the problems that the approaches
discussed here set out to solve. As a result, few domain researchers
may be fully aware of the challenges involved in effectively re-using
published code.

@@

A recent study by researchers at the University of Arizona found that
less than 50% of software could even be successfully built or installed
[@Collberg_2014], and in an ongoing effort by other researchers to
replicate that study [@Krishnamurthi_2014]. Installing or building software
necessary to run the code in question assumes the ability to recreate
the computational environment of the original researchers. 
<!-- Time savings. Personal reproducibility. Collaboration. Teaching. --> 
<!-- Figure 1: technical reproducibility stack
1. Operating system
2. Workflow systems 
3. Package managers 
4. Dynamic documents
-->
<!-- Omit this? -->

Differences in numerical evaluation, such as arise in floating point
arithmetic or computers or even ambiguities in standardized programming
languages ("order-of-evaluation" problems) can be responsible for
differing results between or even within the same computational platform
[@Ince_2012]. Such issues make it difficult to restrict the true dependencies
of the code to higher level environments such as that of a given scripting
language, independent of the underlying OS or even hardware itself.

@@

Documentation on how to install and run code associated with published
research is another frequent barrier to replication. A study by Lapp [@Lapp_2014] found
this impairs a researcher's ability to install and build the software
necessary, as even small holes in the documentation were found to be
major barriers, particularly for "novices" [Sensu @Garijo_2013] --
where novices may be experts in nearby languages but unfamiliar with
the package managers and other tools of the language involved). This
same problem is discussed in [@Clark_2014].  Imprecise documentation
goes well beyond issues of the software environment itself: incomplete
documentation of parameters involved meant as few as 30% of analyses
($n=34$) using the popular software STRUCTURE could be reproduced in the
study of [@Gilbert_2012].

@@

Two dominant paradigms have emerged to address these issues so far:
workflow software [_e.g._ @Altintas_2004; @Hull_2006]  and
cloud computing on virtual machines [_e.g._ @Dudley_2010; @Howe_2012]. 
Workflow software provides very elegant technical solutions to the
challenges of communication between diverse software tools, capturing
provenance in graphically driven interfaces, and handling issues from of
versioning dependencies to data access.  Workflow solutions are often 
built by well-funded collaborations between domain scientists and computer
scientists, and can be very successful in the communities within which 
they receive substantial adoption. Nonetheless, most workflow systems 
struggle with relatively low total adoption overall [@Gil_2007; @Dudley_2010].

@@

<!-- Scripted provisioning: a solution to the weaknesses of virtual machines -->
## A "DevOpts" approach

@@

Sharing the Linux kernel makes Docker much more light-weight and higher
performance than complete virtual machines -- a typical desktop computer
could run no more than a few virtual machines at once but 
would have no trouble running 100's of Docker containers (a container
is simply the term for running instance of an image). This feature
has made Docker particularly attractive to industry and is largely
responsible for the immense popularity of Docker. For our purposes this
is a nice bonus, but the chief value to reproducible research lies
in other aspects.

@@

- It is straight-straight forward for other users to extend or customize 
the resulting image by editing the script directly.

@@

<!-- FIXME move this later --> 
To simplify this process, Docker also supports *Automated Builds*
through the Docker Hub.  This acts as a kind of Continuous Integration
(CI) service that verifies the image builds correctly whenever the
Dockerfile is updated, particularly if the Dockerfile includes checks
for the environment. The Hub also provides a convenient distribution
service, freely storing the pre-built images, along with their metadata,
for download and reuse by others. The Docker Hub is a free service and
an open source software product so that users can run their own private
versions of the Hub on their own servers, for instance, if security of
the data or the longevity of the public platform is a concern.

@@

- Local environment
- Modular reuse
- Portable environments 
- public repository for sharing
- Versioning

@@

While proponents of virtual machines for reproducible research propose
that these machines would be available exclusively as cloud computing
environments [_e.g._ @Dudley_2010], many researchers work _locally_, that
is, primarily with software that is installed on their laptop or desktop
computer, and turning to cloud-based or other remote platforms only for
certain collaborative tasks or when the work is mature enough to need
increased computational power.  Working locally allows a researcher to
rely more on the graphic interface tools for tasks such as managing
files, text editing, debugging, IDEs, or interacting with version
control systems.  Scientific computing on remote machines, by contrast,
still relies largely on potentially less familiar text based command
line functions for these tasks (though web based interfaces are rapidly
filling this gap).

@@

[^2]: External windows, such graphics windows cannot be opened directly
from the container in this setup.  Graphics would have to be saved to
desk as raster or vector files and viewed on the host OS, until a better
solution can be found. RStudio-server is one way around this.

@@

An alternative approach for working locally with familiar tools is to
leverage web-based clients such as RStudio. 
In the R environment, the open source RStudio IDE provides
another key component in making this system accessible to most
domain scientists. Because the RStudio IDE is written completely with
standard web languages (Javascript, CSS, and HTML), its web-based 
RStudio Server looks and feels identical to its popular desktop IDE.
<!--
RStudio Server provides a way to interact with R on a remote environment
without the latency, X tunnelling, etc typically involved in running R
on a remote server. -->

@@

Unlike the much more heavyweight virtual machine approach, containers
are implemented in way such that a single computer can easily run 100s
of such services each in their own container, making it easy to break
computational elements down into logically reusable chunks that come,
batteries included, with everything they need to run reproducibly.
A researcher could connect a container providing a computational 
environment for a different language to this same MySQL container, 
and so forth.

@@

- _Write Dockerfiles instead installing interactive sessions_. As we
have noted already, Docker can be used in a purely interactive manner to
record and distribute changes to a computational environment.  However,
the approach is most useful for reproducible research when researchers
begin by defining their environment explicitly in the DevOps fashion by
writing a Dockerfile.

@@

- Docker is limited to 64 bit host machines, making it impossible to run on
older hardware.

@@

@@

<!--

Upcoming features:

- signed containers?

@@

I noted at the outset that cultural expectations responsible for a
lack of code sharing practices in many fields are a far more extensive
primary barrier to reproducibility than the technical barriers
discussed here. Nevertheless, it may be worth considering how solutions
to these technical barriers can influence the cultural landscape as well.
Many researchers may be reluctant to publish code today because they
fear a it will be primarily a one-way street: more technical savvy
researchers then themselves can benefit from their hard work, while
they may not benefit from the work produced by others.  Lowering the
technical barriers to reuse provides immediate practical benefits
that make this exchanged into a more balanced, two-way street. Another
concern is that the difficulty imposed in preparing code to be shared,
such as providing even semi-adequate documentation or support for other
users to be able to install and run it in the first place is too high
[@Stodden_2010]. Thus, lowering these barrier to re-use through
the appropriate infrastructure may also reduce certain cultural
barriers to sharing.

@@

- *Automatic builds*: Docker defines a format (Dockerfiles) that provide
a simple, scripted way to construct containers, following the DevOpts
philosophy.

>>>>

It is worth observing from the outset that the primary barrier to
computational reproducibility in many domain sciences has nothing
to do with the technological approaches discussed here, but stems rather from a
reluctance to publish the code used in generating the results in the
first place [@Barnes_2010]. Despite extensive evidence to the contrary
[_e.g._ @Ince_2012], many researchers and journals continue to assume that
summary descriptions or pseudo-code provide a sufficient description of
computational methods used in data gathering, processing, simulation,
visualization, or analysis.  Until such code is available in the first
place, one cannot even begin to encounter the problems that the approaches
discussed here set out to solve. As a result, few domain researchers
may be fully aware of the challenges involved in effectively re-using
published code.

@@

A recent study by researchers at the University of Arizona found that
less than 50% of software could even be successfully built or installed
[@Collberg_2014] and similar results are seen in an ongoing effort by other researchers to
replicate that study [@Krishnamurthi_2014]. Installing or building software
necessary to run the code in question assumes the ability to recreate
the computational environment of the original researchers. 
<!-- Time savings. Personal reproducibility. Collaboration. Teaching. --> 
<!-- Figure 1: technical reproducibility stack
1. Operating system
2. Workflow systems 
3. Package managers 
4. Dynamic documents
-->
<!-- Omit this? -->

Differences in numerical evaluation, such as arise in floating point
arithmetic or even ambiguities in standardized programming
languages ("order-of-evaluation" problems) can be responsible for
differing results between or even within the same computational platform
[@Ince_2012]. Such issues make it difficult to restrict the true dependencies
of the code to higher level environments such as that of a given scripting
language, independent of the underlying OS or even hardware itself.

@@

Documentation on how to install and run code associated with published
research is another frequent barrier to replication. A study by Lapp [@Lapp_2014] found
this impairs a researcher's ability to install and build the software
necessary, as even small holes in the documentation were found to be
major barriers, particularly for "novices" [Sensu @Garijo_2013] --
where novices may be experts in nearby languages but unfamiliar with
the package managers and other tools of the language involved. This
same problem is discussed in [@Clark_2014].  Imprecise documentation
goes well beyond issues of the software environment itself: incomplete
documentation of parameters involved meant as few as 30% of analyses
($n=34$) using the popular software STRUCTURE could be reproduced in the
study of [@Gilbert_2012].

@@

Two dominant paradigms have emerged to address these issues so far:
workflow software [_e.g._ @Altintas_2004; @Hull_2006]  and
virtual machines [_e.g._ @Dudley_2010; @Howe_2012]. 
Workflow software provides very elegant technical solutions to the
challenges of communication between diverse software tools, capturing
provenance in graphically driven interfaces, and handling issues from 
versioning dependencies to data access.  Workflow solutions are often 
built by well-funded collaborations between domain scientists and computer
scientists, and can be very successful in the communities within which 
they receive substantial adoption. Nonetheless, most workflow systems 
struggle with relatively low total adoption overall [@Gil_2007; @Dudley_2010].

@@

<!-- Scripted provisioning: a solution to the weaknesses of virtual machines -->
## A "DevOps" approach

@@

Sharing the Linux kernel makes Docker much more light-weight and higher
performing than complete virtual machines -- a typical desktop computer
could run no more than a few virtual machines at once but 
would have no trouble running 100's of Docker containers (a container
is simply the term for running instance of an image). This feature
has made Docker particularly attractive to industry and is largely
responsible for the immense popularity of Docker. For our purposes this
is a nice bonus, but the chief value to reproducible research lies
in other aspects.

@@

- It is straightforward for other users to extend or customize 
the resulting image by editing the script directly.

@@

<!-- FIXME move this later --> 
To simplify this process, Docker also supports *Automated Builds*
through the Docker Hub ([hub.docker.com](http://hub.docker.com)).  This acts as a kind of Continuous Integration
(CI) service that verifies the image builds correctly whenever the
Dockerfile is updated, particularly if the Dockerfile includes checks
for the environment. The Hub also provides a convenient distribution
service, freely storing the pre-built images, along with their metadata,
for download and reuse by others. The Docker Hub is a free service and
an open source software product so that users can run their own private
versions of the Hub on their own servers, for instance, if security of
the data or the longevity of the public platform is a concern.

@@

- Local environment
- Modular reuse
- Portable environments 
- Public repository for sharing
- Versioning

@@

While proponents of virtual machines for reproducible research propose
that these machines would be available exclusively as cloud computing
environments [_e.g._ @Dudley_2010], many researchers work _locally_, that
is, primarily with software that is installed on their laptop or desktop
computer, and turning to cloud-based or other remote platforms only for
certain collaborative tasks or when the work is mature enough to need
increased computational power.  Working locally allows a researcher to
rely more on the graphic interface tools for tasks such as managing
files, text editing, debugging, IDEs, or interacting with version
control systems.  Scientific computing on remote machines, by contrast,
still relies largely on potentially less familiar text based command
line functions for these tasks (though web based interfaces like RStudio
Server are rapidly filling this gap).

@@

[^2]: External windows, such graphics windows cannot be opened directly
from the container in this setup.  Graphics would have to be saved to
disk as raster or vector files and viewed on the host OS, until a better
solution can be found. RStudio Server ([rstudio.com/products/rstudio/#Server](http://rstudio.com/products/rstudio/#Server)) is one way around this.

@@

An alternative approach for working locally with familiar tools is to
leverage web-based clients such as RStudio. 
In the R environment, the open source RStudio IDE provides
another key component in making this system accessible to most
domain scientists. Because the RStudio IDE is written completely with
standard web languages (Javascript, CSS, and HTML), its web-based 
RStudio Server looks and feels identical to its popular desktop IDE.
RStudio Server provides a way to interact with R on a remote environment
without the latency, X tunnelling, and so forth typically involved in running R
on a remote server.

@@

Unlike the much more heavyweight virtual machine approach, containers
are implemented in way such that a single computer can easily run 100s
of such services each in their own container, making it easy to break
computational elements down into logically reusable chunks that come,
batteries included, with everything they need to run reproducibly.
A researcher could connect a container providing a computational 
environment for a different language to this same PostgreSQL container, 
and so forth.

@@

- _Write Dockerfiles instead of installing interactive sessions_. As we
have noted already, Docker can be used in a purely interactive manner to
record and distribute changes to a computational environment.  However,
the approach is most useful for reproducible research when researchers
begin by defining their environment explicitly in the DevOps fashion by
writing a Dockerfile.

@@

- Docker is limited to 64 bit host machines, making it impossible to run on
older hardware (at this time).

@@

- Potential computer security issues may still need to be evaluated. Among other
changes, future support for digitally signing Docker images may make it easier
to build off of only trusted binaries.

@@

@@

I noted at the outset that cultural expectations responsible for a
lack of code sharing practices in many fields are a far more extensive
primary barrier to reproducibility than the technical barriers
discussed here. Nevertheless, it may be worth considering how solutions
to these technical barriers can influence the cultural landscape as well.
Many researchers may be reluctant to publish code today because they
fear a it will be primarily a one-way street: more technical savvy
researchers then themselves can benefit from their hard work, while
they may not benefit from the work produced by others.  Lowering the
technical barriers to reuse provides immediate practical benefits
that make this exchange into a more balanced, two-way street. Another
concern is that the difficulty imposed in preparing code to be shared,
such as providing even semi-adequate documentation or support for other
users to be able to install and run it in the first place is too high
[@Stodden_2010]. Thus, lowering these barriers to re-use through
the appropriate infrastructure may also reduce certain cultural
barriers to sharing.

@@

- *Automatic builds*: Docker defines a format (Dockerfiles) that provide
a simple, scripted way to construct containers, following the DevOps
philosophy.
